\documentclass[a4paper]{exam}

\usepackage{umut}
\usepackage{mathptmx}
\usepackage{uprog}

\usepackage{hyperref}
\hypersetup{colorlinks=true,linkbordercolor=red,linkcolor=green,pdfborderstyle={/S/U/W 1}}

%\printanswers

\pagestyle{headandfoot}
	\lhead{Cogs 502 -- Prog. \& Log. \\ Fall 2017}
	\chead{Homework 2 of 2}
\rhead{Due Jan 15, 2018}
\lfoot{}
\pointname{\%}

\begin{document}

\begin{questions}

\question[50] Use the truth-tree method to determine whether the following formulas are valid or not.

\begin{parts}
\item $p\lor q \cond q\lor p$
\item $((p\imp q) \imp s)\land (p\imp (q\imp s))$
\item $p\lor (q\land r) \cond (p\lor q) \land (p \lor r)$
\item $-(-(-p\land -q)\land -p)\bicond -(q\lor -p)$
\end{parts}


\question[50] Symbolize the following sentences and determine whether they are valid or not. 

\begin{parts}
\item If the three rings have been correctly identified and the mace is indigenous, then the Ajo culture antedated the Tula if and only if the Tula culture was contemporary with or derivative from that of the present excavation. 

\item If it is the case that (if business will flourish on the condition that taxes will not increase, then the standard of living will improve) and (it is not the case that the standard of living will improve) and  (if business does not flourish, then taxes will increase), then unemployment will be a problem.

\end{parts}

\end{questions}

\noindent Submit your solutions to F{\i}rat by the deadline. You may submit your trees as images, but, if you can, please convert your image files to pdf. For this homework, late submissions are not accepted. Please include your name in the name of the file. The homework should be an individual effort. Feel free to consult me or F{\i}rat, if you need further clarification about the homework. But, to avoid ``spoilers'', please do NOT post your queries to the group.

\end{document}
